Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2020-05-15 HRUIonGPU/document.tex
Auto-commit

As shown in Fig.\ \ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig.\ \ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

If the beamformer in (\ref{II_eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{II_eq:b}) and the covariance estimation process in (\ref{II_eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig.\ \ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{II_eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{II_eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{II_eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{II_eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

In \cite{Asen2012, Buskenes2015} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{II_fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{II_eq:bs_formula}).
This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
%Equation (\ref{II_eq:sliding_bs}) shows that 
The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity. In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Fig.\ \ref{II_fig:gpulayoutbs}, is to combine SDFT with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

\ifPhdDoc
  \begin{figure*}[!t]
  \centerline{\subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
%   \hfil
  \subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
  \label{II_fig:benchBS96}}}
\else
  \begin{figure*}[!t]
  \centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
  \hfil
  \subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
  \label{II_fig:benchBS96}}}
\fi%
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Fig.\ \ref{II_fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{II_fig:benchBS}
\end{figure*}
%
In this section we present benchmarks of the implementations presented in Sections \ref{II_sec:meth} and \ref{II_sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
%
\begin{equation}\label{II_eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
%
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

@@

Benchmarks of the ES-Capon beamformer are presented in Fig.\ \ref{II_fig:bench}. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{II_sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS). For $M=96$, $L=48$ and $K=2$ the processing throughput drops below $0.5$ FPS.

In Fig.\ \ref{II_fig:benchBS} we present similar benchmarks for the BS-Capon beamformer. The figure only presents results with use of temporal averaging ($K=2$). For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 FPS (8 ms). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 FPS (13 ms). In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit. Hence, below this line we are capable of performing real-time processing.

@@

%\begin{figure*}[!t]
%\centerline{
%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{II_fig:simulationSlice}}
%\hfil
%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{II_fig:invivoSlice}}
%}
%\caption{Lateral slices taken from the images in Fig. \ref{II_fig:phantom} and Fig. \ref{II_fig:invivo}. a) From Fig. \ref{II_fig:phantom} at 93 mm constant depth, through the two bright point scatterers. b) From Fig. \ref{II_fig:invivo} at 96 mm range, through the tip of the mitral valve.}
%\label{II_fig:slices}
%\end{figure*}

%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Fig. \ref{II_fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

Fig.\ \ref{II_fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle. The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive). The image frame is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 1, 2 and 3}.
The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{II_eq:rayleigh}).
The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.  %It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.

%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Fig. \ref{II_fig:phantom} and \ref{II_fig:invivo} are included in Fig. \ref{II_fig:slices}. For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.

@@

\section{Trading resolution for speed}\label{II_sec:trade}
As seen from Fig.\ \ref{II_fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced. The same is true when $N_b$ is reduced in beamspace. So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace. In this section we investigate how adjusting $L$ for ES-Capon and $N_b$ for BS-Capon impacts the lateral resolution.

In Fig.\ \ref{II_fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon). To produce the result in Fig.\ \ref{II_fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$. 
In Fig.\ \ref{II_fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$. %The resolution for ES-Capon is therefore better here than what is presented in Fig. \ref{II_fig:phantom}. 
%We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf. Section \ref{II_sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Fig. \ref{II_fig:phantom}.

@@

\begin{figure}
\centering
\includegraphics[width=3.3in]{gfx/speed_res_trade.png}
\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension. As $L$ or $N_b$ gets smaller, the execution time is reduced (Fig.\ \ref{II_fig:bench} and \ref{II_fig:benchBS}). When $N_b$ is varied, $L$ is fixed on 32.}
\label{II_fig:speed_res_trade}
\end{figure}

\section{Discussion}\label{II_sec:dis}
In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU. According to Fig.\ \ref{II_fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$. $L=M/2$). %However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate. If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling. If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.

For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging. From Fig.\ \ref{II_fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS). Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al. \cite{Chen2011}.

A major part of the total execution time for ES-Capon comes from solving the system of linear equations. This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Fig.\ \ref{II_fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible. The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time. If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Fig.\ \ref{II_fig:benchUS64} and \ref{II_fig:benchBS} are real-time. More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image. With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 FPS respectively using single line acquisition. Each benchmark plot has a line indicating this real-time requirement. With beamspace processing, all processing times for large subarrays are below this line.

@@

%Turning to image quality we see from Fig. \ref{II_fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$). Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}. %How large this oversampling factor needs to be and how the oversampling in the end is conducted will impact the real time requirement, and . 
%Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
%No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.

>>>>

As shown in Figure~\ref{II_fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. As with coarse-fine beamformers \cite{Thomenius}, the complex weights cause a micro-steering which is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Figure~\ref{II_fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of executing multiple arithmetic operations per clock cycle.
%32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

@@

If the beamformer in (\ref{II_eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{II_eq:b}) and the covariance estimation process in (\ref{II_eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Figure~\ref{II_fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{II_eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{II_eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{II_eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{II_eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance in cardiac ultrasound imaging can only be achieved by exploring methods with reduced complexity.

@@

In \cite{Asen2012, Buskenes2015} we demonstrated how the cubic complexity with $L$ %and linear complexity with $N_K$ 
can be reduced by realizing that calculations in (\ref{II_eq:R}) overlap across subarrays. %Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
%\begin{align}
%\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
%&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
%&- x_{i-1}[n']x_{j-1}[n']^H). \label{II_eq:sliding}
%\end{align}
%Along the time dimension we also find that
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
%&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. 
%In that case, 
The complexity of calculating $\mat{\hat{R}}$ is by this reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.
%The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Figure~\ref{II_fig:gpulayout}). 
Maximum supported $L$ is also increased to the maximum number of threads per block (currently 1024). %However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

%We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{II_eq:bs_formula}).
This formula has been implemented using a normalized sliding DFT (SDFT)\cite{Lyons2003}.
%Equation (\ref{II_eq:sliding_bs}) shows that 
The SDFT allows us to still save instructions by utilizing the overlap between subarrays, but with an increase in task level granularity. In this situation, it might be tempting to use one thread per beam (index $k$) that calculates this beam for all subarrays. However, since the number of selected beams usually is small ($N_b \ll M$) we will end up with a large memory-per-thread ratio ($M/N_b$). Using one thread per beam is therefore not feasible for small-$N_b$-large-$M$ configurations. One solution, as depicted in Figure~\ref{II_fig:gpulayoutbs}, is to combine SDFT with full computations. A number of sub thread groups of size $N_b$ is created that targets different sets of subarrays which they transform using the SDFT algorithm. In that way, the number of threads per sample is increased and thus the memory-per-thread ratio is decreased.

@@

\ifPhdDoc
  \begin{figure*}[!t]
  \centerline{\subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
%   \hfil
  \subfloat[]{\includegraphics[width=3.8in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
  \label{II_fig:benchBS96}}}
\else
  \begin{figure*}[!t]
  \centerline{\subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=64_Nx=80_Ny=540_Nb=3.png} \label{II_fig:benchBS64}}
  \hfil
  \subfloat[]{\includegraphics[width=3.4in]{gfx/benchmark_bar_bs_M=96_Nx=120_Ny=540_Nb=3.png}%
  \label{II_fig:benchBS96}}}
\fi%
\caption{Benchmarks of GPU beamspace Capon beamforming for a cardiac image covering a $70 \degree$ and $15.4$ cm deep sector. Execution time is plotted as a function of subarray length $L$.  If $L$ had been fixed at $M/2$ and execution time was plotted against $N_b$, the solver would have shown the same quadratic growth as in Figure~\ref{II_fig:bench}. When $N_b$ is fixed, calculating the covariance matrix and the BS-transform is inverse proportional and proportional to $L$ respectively. The large jumps in execution times are caused by adaptive kernel launch configurations, which depends on $M$, $L$ and $N_b$.}
\label{II_fig:benchBS}
\end{figure*}
%
In this section we present benchmarks of the implementations presented in Sections \ref{II_sec:meth} and \ref{II_sec:bs} performed on a Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops theoretical single precision performance, 6 GB global memory, and 144 GB/s memory bandwidth). Problem sizes are calculated based on 64 and 96 element phased arrays with a 3.4 MHz center frequency. Assuming a system bandwidth of 80\% (relative to the center frequency) we need a complex sampling frequency of 2.7 MHz, and when imaging down to a depth of 15.4 cm we have approximately 540 samples per range line. The angular sampling is given by the Rayleigh criterion divided by two \cite{Hergum2007}, 
%
\begin{equation}\label{II_eq:rayleigh}
\delta{}\theta = \lambda/(2Mp),
\end{equation}
%
where $\lambda$ is the pulse wavelength and $p$ is the element pitch. 
To sample a $70\degree$ sector with $64$- and $96$-element arrays with $p=\lambda/2$ we therefore need approximately 80 and 120 receive beams respectively.

@@

Benchmarks of the ES-Capon beamformer are presented in Figure~\ref{II_fig:bench}. For each subarray size, the total execution time is split into one bar for each of the three steps described in Section \ref{II_sec:meth}, and the calculation of $\mat{\hat{R}}$ is presented for three different values of $K$. For $M=64$, $L=32$ and $K=2$, a typical setup for high-quality cardiac imaging, processing throughput is 8.3 frames per second (FPS). For $M=96$, $L=48$ and $K=2$ the processing throughput drops below $0.5$ FPS.

In Figure~\ref{II_fig:benchBS} we present similar benchmarks for the BS-Capon beamformer. The figure only presents results with use of temporal averaging ($K=2$). For $M=64$, $L=32$, $K=2$ and $N_b=3$, processing throughput is 125 FPS (8 ms). For $M=96$, $L=48$, $K=2$ and $N_b=3$, processing throughput is 77 FPS (13 ms). In all four benchmarks there is a line indicating the time it takes to acquire the underlying image using one receive line per transmit. Hence, below this line we are capable of performing real-time processing.

@@

%\begin{figure*}[!t]
%\centerline{
%\subfloat[]{\includegraphics[width=3in]{gfx/simulation_slice.png} \label{II_fig:simulationSlice}}
%\hfil
%\subfloat[]{\includegraphics[width=3in]{gfx/invivo_slice.png} \label{II_fig:invivoSlice}}
%}
%\caption{Lateral slices taken from the images in Figure~\ref{II_fig:phantom} and Figure~\ref{II_fig:invivo}. a) From Figure~\ref{II_fig:phantom} at 93 mm constant depth, through the two bright point scatterers. b) From Figure~\ref{II_fig:invivo} at 96 mm range, through the tip of the mitral valve.}
%\label{II_fig:slices}
%\end{figure*}

%To investigate the image quality of both the ES and BS-Capon beamformer, a phantom was simulated using Field II \cite{Jensen1992, Jensen1996} (Figure~\ref{II_fig:phantom}). The simulation was performed using a 64-element, $p=\lambda/2$ pitch, 10 $\mu$m kerf, 80\% bandwidth, and 10 mm high phased array. The transducer was excited with a 3.4 MHz and $1.5\lambda$ long pulse with transmit focus at 90 mm. The phantom contains two circular areas, one bright at 20 dB and one dark at -20 dB, and two point targets at 30 dB, located at 90 mm depth. Speckle level is at 0 dB, covering an area from $\pm90\degree$ in azimuth. The resulting image has been processed both using DAS with uniform weights, ES- and BS-Capon. The presented image is frame $15$ in a series of images where motion is imposed on both circles, point targets and the probe. The entire simulation is presented in two attached videos \multimedia{Media-Movie 1 and 2}.

Figure~\ref{II_fig:invivo} presents results of applying ES- and BS-Capon on an \textit{in-vivo} image of the left ventricle. The image was acquired using a 64-element and $p=0.65\lambda$ pitched phased array in harmonic mode (1.7 MHz at transmit and 3.4 MHz at receive). The image frame is extracted from a cardiac cycle, that can be viewed in its full length in the attached videos \multimedia{Media-Movie 1, 2 and 3}.
The region of interest is 70\degree{} with a minimal lateral sampling of 98 receive beams given by (\ref{II_eq:rayleigh}).
The image was acquired using 50 transmit beams and "Synthetic Transmit Beamforming" was used on receive \cite{Hergum2007}.  %It is included to show that both ES-Capon and BS-Capon beamforming are feasible for \textit{in-vivo} cardiac imaging, not just in simulations.

%To get a better impression of the improved lateral resolution obtained with Capon beamforming, two slices taken from the images in Figure~\ref{II_fig:phantom} and \ref{II_fig:invivo} are included in Figure~\ref{II_fig:slices}. For the phantom, the cut goes through the two point targets at 90 mm constant radius, and for the cardiac image it goes through the tip of the mitral valve at 96 mm constant radius.

@@

\section{Trading resolution for speed}\label{II_sec:trade}
As seen from Figure~\ref{II_fig:bench} the execution time, for the solver in particular, is reduced when $L$ is reduced. The same is true when $N_b$ is reduced in beamspace. So to gain an  increase in processing throughput we can reduce $L$ in element space or $N_b$ in beamspace. In this section we investigate how adjusting $L$ for ES-Capon and $N_b$ for BS-Capon impacts the lateral resolution.

In Figure~\ref{II_fig:speed_res_trade}, the minimum distance between two point targets (while preserving a 6 dB saddle point in between the two) is plotted against $L$ (when processed with ES-Capon) and $N_b$ (when processed with BS-Capon). To produce the result in Figure~\ref{II_fig:speed_res_trade} $K=2$ and $d=0.01$ has been used on both methods in order to obtain the same resolution for $L=N_b=M/2$. 
In Figure~\ref{II_fig:invivo}, the diagonal loading factor for BS-Capon was adjusted to make the two results equal for $L=32$ and $N_b=3$. %The resolution for ES-Capon is therefore better here than what is presented in Figure~\ref{II_fig:phantom}. 
%We considered this to be a small value for ES-Capon and a normal value for BS-Capon (cf. Section \ref{II_sec:bs}), hence the resolution for ES-Capon is here better than what is presented in Figure~\ref{II_fig:phantom}.

@@

\begin{figure}
\centering
\includegraphics[width=3.3in]{gfx/speed_res_trade.png}
\caption{Trade-off between resolution and speed by reducing subarray size or reducing beamspace dimension. As $L$ or $N_b$ gets smaller, the execution time is reduced (Figure~\ref{II_fig:bench} and \ref{II_fig:benchBS}). When $N_b$ is varied, $L$ is fixed on 32.}
\label{II_fig:speed_res_trade}
\end{figure}

\section{Discussion}\label{II_sec:dis}
In this paper we have presented a parallel implementation of the BS-Capon beamformer for high-resolution cardiac ultrasound imaging. The results show that it is now possible to apply BS-Capon in real-time in cardiac ultrasound imaging by utilizing the power of a GPU. According to Figure~\ref{II_fig:benchBS} it is also possible to process two parallel receive lines per transmit beam (for both $M=64$ and $M=96$. $L=M/2$). %However, since we had to use oversampling in our simulation, these extra lines might be needed for increased sampling instead of increasing the frame rate. If parallel receive lines are used to obtain denser sampling, the acquisition time remains the same but the processing time is increased with a factor equal to the amount of oversampling. If parallel receive lines are used to increase the frame rate, the acquisition time is lowered, and the processing time remains constant. Hence, four parallel receive lines are not possible to process in real-time using BS-Capon with $N_b=3$ in combination with our single GPU setup.

For ES-Capon, processing rates for high quality settings are still too high to support real-time cardiac ultrasound imaging. From Figure~\ref{II_fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, the total execution time is 110 ms (9 FPS). Taking into account that we use complex data, but less samples per image, these results are similar to the performance reported by Chen et al. \cite{Chen2011}.

A major part of the total execution time for ES-Capon comes from solving the system of linear equations. This limitation is targeted by reducing the sample covariance matrix size in beamspace, and as shown in Figure~\ref{II_fig:benchBS}, inverting the $3\times3$ beamspace covariance matrix is then barely visible. The time it takes to construct all covariance matrices is also reduced, but for few beams this is now the major contributor to the total execution time. If real-time is defined as 10 FPS \cite{Chen2011} almost all execution times in both Figure~\ref{II_fig:benchUS64} and \ref{II_fig:benchBS} are real-time. More natural for cardiac ultrasound imaging is to define real-time as the time it takes to acquire the underlying image. With this requirement the images formed using a 64 and 96 element array have a real-time requirement of 62.5 and 41.7 FPS respectively using single line acquisition. Each benchmark plot has a line indicating this real-time requirement. With beamspace processing, all processing times for large subarrays are below this line.

@@

%Turning to image quality we see from Figure~\ref{II_fig:phantom}, and the attached video of the simulation, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$). Our preliminary finding is that the Capon beamformers are local shift invariant as long as the lateral sampling is increased to a certain level in simulations \multimedia{Media-Movie 1 and 2}. %How large this oversampling factor needs to be and how the oversampling in the end is conducted will impact the real time requirement, and . 
%Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
%No major improvement in contrast is visually observed in the presented images and videos, other than slightly rounder circles due to mainlobe micro steering.

Turning to image quality we see from Figure~\ref{II_fig:speed_res_trade} and the attached videos, that Capon beamforming provides improved lateral resolution of moving point targets both in element- and in a reduced beam-space ($N_b=3$). In the cardiac recording (Figure~\ref{II_fig:invivo}), structures are thinner and better resolved when the Capon beamformers are applied. One example is the mitral valve located at 10 cm depth and -1.5 cm width. On the other hand we observe a loss in contrast, especially in the septum (Figure~\ref{II_fig:invivo}, 6 cm depth, -2 cm width). This loss can be partially explained by the presence of highly correlated signals and mismatch between the assumed steering vector and signal. This is the same issue that made us use 16x oversampling when producing Figure~\ref{II_fig:speed_res_trade}. Future work should focus on how large this oversampling needs to be and how it should be conducted in order to still achieve real time processing.
